---
title: State of Crossref metadata by member country
date: 2026-05-24
author: 
    name: Luis Montilla
    orcid: 0000-0002-7079-6775
categories:
- TidyTuesday
- Scholarly metadata
- R
- Dissertations
execute:
  echo: false
  warning: false
  freeze: auto
format:
    html:
        code-tools: true
---

```{r}
#| label: libraries
#| warning: false
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
library(ggrepel)
library(ggiraph)
library(ggpomological)
```

```{r}
#| label: member_participation_stats_by_country
raw_participation <- read_csv("https://zenodo.org/api/records/19928426/files/member_participation_stats_by_country.csv/content")

raw_coverage <- readr::read_csv("https://zenodo.org/api/records/19928426/files/metadata_coverage_stats_by_country.csv/content")
```

I'm returning from by partcipation at the EuroCRIS 2026 conferece thiking a lot about dissertations and other content types, so I thought it was appropriate to explore the metadata of this content type using Crossref data in the context of Tidy Tuesday[^1]. For that I'll filter data out to keep only the past quarter (Q1 2026) and only the content type `dissertation`.

In the following plot let's explore which region accumulates more DOI deposits and out of those, what is the percetage of dissertations:

```{r}
total_records <- raw_coverage |>
    filter(current_up_to == "2026-03-31") |>   # Q1 2026
        summarise(
            total_records = sum(n_dois), 
            .by = region_id
            )
```

```{r}
#| echo: false
df <- raw_coverage |> 
        filter(current_up_to == "2026-03-31" & document_type == "dissertation") |>   # Q1 2026
        summarise(
            dissertations = sum(n_dois), 
            with_ref = sum(with_ref),
            with_abs = sum(with_abstract),
            with_orc = sum(with_orcid),
            with_ror_aff = sum(with_ror_id_for_affiliations),
            with_epp = sum(with_crossmark),
            .by = region_id
            ) |>
        left_join(total_records) |>
        mutate(
            label = case_when(
                region_id == "EAS" ~ "East Asia & Pacific",
                region_id == "ECS" ~ "Europe & Central Asia",
                region_id == "LCN" ~ "Latin America & the Caribbean",
                region_id == "MEA" ~ "Middle East, North Africa, Afghanistan & Pakistan",
                region_id == "NAC" ~ "North America",
                region_id == "SAS" ~ "South Asia",
                region_id == "SSF" ~ "Sub-Saharan Africa", 
                .default = NULL
                )
            ) |> 
        select(-region_id)
```

```{r}
props_region <- df |> 
    mutate(
        across(where(is.double) & !(total_records), ~ ((./total_records)*100))
        ) |>
    ggplot(aes(
            x = total_records, 
            y = dissertations
            )
        ) +
        geom_point_interactive(aes(
            tooltip = total_records, 
            data_id = label
        )) +
        geom_text_repel(aes(
        label = label,
        family = "Inter"
    )) +
        labs(
            y = "Percentage of dissertations (%)",
            x = "Number of DOI records"
        ) +
        theme_minimal()
 
girafe(props_region)
```

Interestingly, Latin America & the Caribbean clearly have deposited more diisertations records than other regions.

One important aspect of the conversation that is not always evident, is that we should also take into account the quality and completeness of the metadata. 

We could have 100% of an institution's production assigned with DOIs and those could also be just persistent IDs, without fully realizing the full potential of the potential of this tool. 

Considering this, let's explore out of a region's total dissertation records, what's the percetage of records that includes some of the Crossref recommended metadata[^2]:

```{r}
df |> 
    mutate(
        across(starts_with("with_"), ~ (round((./dissertations)*100, 2)))
        ) |> 
    select(-c(total_records, dissertations)) |> 
    gt(rowname_col = "label") |>
        data_color(
            palette = "viridis",
            alpha = 0.3,
            autocolor_text = FALSE
            ) |> 
        cols_label(
            with_ref = "References",
            with_abs = "Abstracts",
            with_orc = "ORCIDs",
            with_ror_aff = "ROR IDs",
            with_epp = "Policy pages"
        ) |> 
            opt_interactive(
                use_pagination = FALSE,
                use_compact_mode = TRUE  
            )
```

As expected, there are complexities in the scholarly record. From the first plot we see that Latin America & the Caribbean is the region that accumulates more dissertations, but each region has paid more attention to specific metadata fields. E.g. Sub-Saharan Africa leads the way with the percentage of dissertations metadata that include references and abstracts.

[^1]: [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday)
[^2]: [Crossref documentation: Required, recommended, and optional metadata](https://www.crossref.org/documentation/schema-library/required-recommended-elements/)